---
title: "Exercise 4.1"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

a. $P(Z < 1.33)$

    This is a left tailed probability, and can be calculated directly in R:

    ```{r}
    pnorm(1.33)
    ```

b. $P(Z > 1.33)$

    This is a right tailed probability, but can be converted to a left tailed probability using the 100% rule.

    ```{r}
    1-pnorm(1.33)
    ```

c. $P(Z < -1.33)$

    This is a left tailed probability, and can be calculated directly in R:

    ```{r}
    pnorm(-1.33)
    ```
    
    We get the same result as in question b., which is expected because of symmetry.
  

d.  $P(Z > -1.33)$

    This is a right tailed probability, but can be converted to a left tailed probability using the 100% rule.

    ```{r}
    1-pnorm(-1.33)
    ```
    
    We get the same result as in question 1., which is expected because of symmetry.

e. $P(Z < 0.45)$

    This is a left tailed probability, and can be calculated directly in R:

    ```{r}
    pnorm(.45)
    ```

f. $P(Z > -1.05)$

    This is a right tailed probability, but can be converted to a left tailed probability using the 100% rule.

    ```{r}
    1 - pnorm(-1.05)
    ```
    
g. $P(Z < 0.65)$

    ```{r}
    pnorm(.65) 
    ```

h. $P(-0.45 < Z < 1.20)$

    We can calculate this probability by taking the difference of 2 left tailed probabilities:
    
    $P(-0.45 < Z < 1.20) = P(Z < 1.20) - P(Z < -0.45)$ 
    
    ```{r}
    pnorm(1.2) - pnorm(-.45)
    ```
    
i. $P(-1.35 < Z < -0.10)$

    Similar to previous question
    
    ```{r}
    pnorm(-.1) - pnorm(-1.35)
    ```
    
j. $P(-2.10 < Z < -0.90)$

    Similar to previous question
    
    ```{r}
    pnorm (-.9) - pnorm(-2.1)
    ```
