---
title: "Exercise 4.4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Solution 4.4

For this exercise you will need the package ggplot2 Information can be found [here](https://www.dummies.com/programming/r/how-to-install-and-load-ggplot2-in-r/)


## Part 1

Use the `rnorm()` function to generate a random sample of 25 values based on a normal distribution with mean $\mu = 0$ and standard deviation $\sigma = 1$.

```{r }

my_theme <- theme(plot.title = element_text(family = "Helvetica", face = "bold", size = (15)), 
                 legend.title = element_text(colour = "steelblue",  face = "bold.italic", family = "Helvetica"), 
                 legend.text = element_text(face = "italic", colour="steelblue4",family = "Helvetica"), 
                  axis.title = element_text(family = "Helvetica", size = (10), colour = "steelblue4"),
                  axis.text = element_text(family = "Courier", colour = "cornflowerblue", size = (10)))
  


overlayplot <- function(n,m,s){
  dataset <- tibble(obs = rnorm(n,m,s))
  ggplot(data = dataset,mapping = aes(x=obs)) + 
    geom_histogram(aes(y = ..density.. ), alpha = 0.7, fill = "#333333") + 
    labs(y="observed", x="observations", title = "Histogram")+
    geom_density(fill = "#ff4d4d",alpha = 0.5) + 
    stat_function(fun = dnorm, args = list(mean = m, sd = s),
                  colour = "red")+
    my_theme
}

overlayplot(25,0,1)

```

## Part 2 

Draw an overlay on the histogram consisting of:

- (a) the theoretical probability distribution curve of a normal distribution with mean $\mu = 0$ and standard deviation $\sigma = 1$
- (b) an ``estimated'' probability distribution curve based on the measured sample mean and standard deviation.

### 100
```{r }
overlayplot(100,0,1)
```


### 500
```{r }
overlayplot(500,0,1)
```


### 1000
```{r }
overlayplot(1000,0,1)
```

### 10000
```{r }
overlayplot(10000,0,1)
```
