---
title: "Exercise 4.5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Exercise 4.5
The course of research techniques at a college is organised in two separate classes. 
Students were distributed at random over the two classes, A and B, so we can assume that one class is not significantly better than the other. 
Class A is lectured by Mrs. X, class B by Mr. Y.

Mrs. X is rather strict and her group gets grades with an average of 54/100 and a standard deviation of 11.
Y is less strict and is known for stimulating his students by giving them a bonus point at times. By the end of the academic year, the grades for his group have an average of 62/100 with a standard deviation of 7.

Wouter is a student of class A and got a grade of $\frac{63}{100}$, Stijn from class B got $\frac{67}{100}$. Who has the best result?

## Solution

At first sight, it looks like Stijn has a better grade, as he has more points than Wouter.
But is this correct? We have to investigate how well both students performed, compared to the rest of their class.
In other words, how extreme are the values of Wouter and Stijn compared to the mean of the class.
This can be done by standardizing: we have to calculate the $z$-score for both values and compare them.

\[ Z = \frac{x - \mu}{\sigma} \]

```{r cars}
score_wouter = 63
score_stijn = 67
m_wouter = 54
m_stijn = 62
s_wouter = 11
s_stijn = 7

normalisation <- function(x,m,s){
  y <- (x-m)/s
  return (y)
}

z_wouter = normalisation(score_wouter,m_wouter,s_wouter)
z_stijn = normalisation(score_stijn, m_stijn,s_stijn)

z_wouter
z_stijn

perc_wouter <- 1-z_wouter
perc_stijn <- 1 -z_stijn
```

We have that the standardized score of Wouter is larger than the score of Stijn, and therefore Wouter did better than Stijn, although Stijn had a higher grade. Furthermore:

- Wouter is in the top `r 100*round(perc_wouter,2)` % of the class
- Stijn in in the top `r 100*round(perc_stijn,2)` % of the class

We can also illustrate this by plotting graphs:

```{r }
plotNormal <- function(x,m,s){
  # Area left of x
xvals <- seq(from = m-4*s,
             to = m+4*s,
             length.out = 201)
i <- xvals >= x

ggplot(mapping = aes(x = xvals)) +
  # Show the area left of x and below the Gauss-curve in green
  geom_area(mapping = aes(x = xvals[i], y = dnorm(xvals[i], m, s)), 
            color = "black", fill = "cornflowerblue") +
  # Plot the Gauss-curve
  stat_function(fun = dnorm, args = list(mean = m, sd = s)) +
  # Draw the mean using a red vertical line
  geom_vline(xintercept = m, color = "red") +
  # Adjust the axis labels
  xlab("x") + ylab("densiteit") +
  # Add text to the plot
  annotate("text", x = x, y = .02, label = x) +
  annotate("text", x = m+.1, y = .02, label = m, color = "red")
}

plotNormal(score_wouter,m_wouter,s_wouter)
```

## Wouter
```{r }
plotNormal(score_wouter,m_wouter,s_wouter)
plotNormal(score_stijn,m_stijn,s_stijn)
```
