---
title: "Exercise 4.6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

> A survey conducted between 1988 and 1994 indicated that the average cholesterol level of women between 20 and 29 years was 183 mg/dl, with a standard deviation of 36. The sample consisted of 81 women, and we assume that the sample was taken at random.

```{r}
m <- 183
s <- 36
n <- 81
```

# probability distribution

> Plot the probability distribution for the sample mean $\overline{x}$.

```{r}
sn <- s/sqrt(n)
ggplot(data = data.frame(x = c(m-4*sn, m+4*sn)), mapping = aes(x)) +
  stat_function(fun = dnorm, n = 201, args = list(mean = m, sd = sn)) +
  geom_vline(xintercept = m, color = "green")
```

# Calculating probabilities

> What is the probability that $\overline{x}$ is smaller than 185?

```{r}
pnorm(185, mean = m, sd = s / sqrt(n))
```

> What is the probability that $\overline{x}$ is between 175 and 185?

```{r}
pnorm(185, mean = m, sd = s / sqrt(n)) - pnorm(175, mean = m, sd = s / sqrt(n))
```

> What is the probability that $\overline{x}$ is larger than 190? 

```{r}
1 - pnorm(190, mean = m, sd = s / sqrt(n))
```

