---
title: "Oefening 4.7"
author: "Bert Van Vreckem"
date: "20/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

> A random sample consisting of 64 observations is drawn from a population with an unknown distribution. However, the population mean and standard deviation are known: $\mu = 20$ and $\sigma = 16$.

```{r}
m <- 20
s <- 16
n <- 64

```

# Probability distribution sample mean

> What are the mean and standard deviation of the sample?

- Expected value for population mean: $\mu = 20$
- Standard deviation: $\sigma/\sqrt{n} = 16/\sqrt{64} = 2$

```{r}
sn <- s / sqrt(n)
```

> Describe the shape of the probability distribution of the sample mean. Does this depend on the sample size?

\[ M \sim Nor(\mu = 20, \sigma = 16/\sqrt{64}))$ \]

```{r}
ggplot(data = data.frame(x = c(m-4*sn, m+4*sn)), mapping = aes(x)) +
  stat_function(fun = dnorm, n = 201, args = list(mean = m, sd = sn)) +
  geom_vline(xintercept = m, color = "green")
```


# z-score

> Calculate the $z$-score for $\overline{x_{1}} = 15.5$ and $\overline{x_{2}} = 23$.

```{r}
(15.5 - m) / sn
(23 - m) / sn
```

# Calculating probabilities

> Calculate the probability that $\overline{x} < 16$.

```{r}
pnorm(16, mean = m, sd = sn)
```

> Calculate the probability that $\overline{x} > 23$.

```{r}
1 - pnorm(23, mean = m, sd = sn)
```

> Calculate the probability that $16 < \overline{x} < 22$.

```{r}
pnorm(22, mean = m, sd = sn) - pnorm(16, mean = m, sd = sn)
```
