---
title: "Exercise 4.9"
output: html_document
---

Lately, a canned goods manufacturer received complaints about the net contents of their carrot and peas cans.
According to the label, the can should contain 1 liter.  
To verify these complaints, the QA department draws a random sample of 40 cans and checks the contents. 
The results are provided in the table below.


| Contents      | $n_i$ |
| :---          | :---  |
| $[970,980[$   | 3     |
| $[980,990[$   | 5     |
| $[990,1000[$  | 13    |
| $[1000,1010[$ | 11    |
| $[1010,1020[$ | 5     |
| $[1020,1030[$ | 3     |
    
## Question A:


- Add the cumulative absolute frequency to the table

    We can calculate this using the `cumsum` function:

    ```{r}
    n_i <- c(3,5,13,11,5,3)         # Given frequencies
    cf  <- c(cumsum(n_i))           # Cumulative absolute Frequencies
    cf
    ```
    
- Add the relative frequencies

    The relative frequency is the absolute frequency, normalised by the total amount.
    $f_i = n_i/N$

    In this example, $N = 40$ or the sample size. 
   
    
    ```{r}
    f_i <- n_i / sum(n_i)
    f_i
    ```

- Add the cumulative relative frequencies

    These can also be calculated using the `cumsum` function
    
    ```{r}
    crf <- c(cumsum(f_i))
    crf
    ```
    
    Note that the last value needs to be 1! (why?)
    
Full table:
    

| Contents      | $n_i$ | CF   | $f_i$ | CRF   |  
| :---          | :---  | :--- | :---  | :---  |
| $[970,980[$   | 3     | 3    | 0.075 | 0.075 |
| $[980,990[$   | 5     | 8    | 0.125 | 0.200 |
| $[990,1000[$  | 13    | 21   | 0.325 | 0.525 |
| $[1000,1010[$ | 11    | 32   | 0.275 | 0.800 |
| $[1010,1020[$ | 5     | 37   | 0.125 | 0.925 |
| $[1020,1030[$ | 3     | 40   | 0.075 | 1.000 |

In R:

```{r}
conserven <- tibble(
  category = c("[970,980[","[980,990[","[990,1000[","[1000,1010[","[1010,1020[","[1020,1030["),
  n_i = c(3,5,13,11,5,3),
  cf  = cumsum(n_i),
  f_i = n_i / sum(n_i),
  crf = cumsum(f_i)
)
```

## Question B:

- Calculate the mean

    Cfr. Exercise 3.07. For the x values we take the mean of the interval.
    
    ```{r}
    x <- c(975,985,995,1005,1015,1025)
    m <- sum(x * n_i) / sum(n_i)
    m
    ```

- Calculate the standard deviation

    Cfr. Exercise 3.07.
    
    ```{r}
    v <- sum(n_i * (x - m)^2) / (sum(n_i) - 1)
    stdev <- sqrt(v)
    stdev
    ```
    
- What percentage of cans does not contain sufficient carrots and peas?

    There are 3 cans with contents $[970,980[$, 5 cans with $[980,990[$ and 13 cans with contents $[990,1000[$.
    There are 21 with not insufficient contents, this corresponds to the 3rd value of the cumulative absolute frequencies.
    
    ```{r}
    cf[3]/sum(n_i)
    ```
    
    $52.5\%$ of the cans is not sufficiently filled.
    
- Plot a histogram of the absolute frequency

  ```{r}
  names(n_i) <- c("[970,980[","[980,990[","[990,1000[","[1000,1010[","[1010,1020[","[1020,1030[")
  barplot(n_i)
  ```