---
title: "Exercise 4.11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Exercise 4.11 - Confidence intervals for "money.csv"

Import the dataset `money.csv` in R. We assume that the values of this sample are normally distributed with 
  an unknown population mean $\mu$, but the standard deviation of the population
  is supposed to be known; $\sigma = 99$. 

```{r}
money <- read_csv("../../../onderzoekstechnieken-cursus/oefeningen/datasets/money.csv")

pop_sd <- 99  # standard deviation of the population (supposed to be known)

n <- length(money$Money)
sample_mean <- mean(money$Money) # sample mean
sample_sd <- sd(money$Money)     # standard deviation of the sample
```

## 99% confidence interval

> Calculate a 99\% confidence interval for the population mean

```{r}
z_99 <- qnorm(1 - 0.01/2)

sprintf("99%% CI : [%.3f,%.3f]",
        sample_mean - z_99 * pop_sd / sqrt(n),
        sample_mean + z_99 * pop_sd / sqrt(n))
```

## 95% confidence interval

> Calculate a 95\% confidence interval for the population mean

```{r}
z_95 <- qnorm(1 - 0.05/2)

sprintf("95%% CI : [%.3f,%.3f]",
  sample_mean - z_95 * pop_sd / sqrt(n),
  sample_mean + z_95 * pop_sd / sqrt(n))

```

## Unknown σ

> Suppose that $\sigma$ is unknown. Calculate a 95\% confidendce interval for the population mean.

If $\sigma$ is unknown, we cannot use the z-score, but have to use the t-score instead.

```{r}
t_95 <- qt(1 - 0.05/2, df = n-1)

sprintf("Unknown σ, 95%% CI : [%.3f,%.3f]",
        sample_mean - t_95 * sample_sd / sqrt(n),
        sample_mean + t_95 * sample_sd / sqrt(n))
```

## Small sample

> Finally, calculate a 95% confidence interval for the population mean, assuming that the sample only consists of the first 25 values of the file.

For a small sample we also need to use the t-score.

```{r}
small_sample <- money$Money[1:25]
small_sample_mean <- mean(small_sample)
small_sample_sd <- sd(small_sample)
small_sample_n <- length(small_sample)

t_95 <- qt(1 - 0.05/2, df = small_sample_n - 1)

sprintf("Small sample, 95%% CI : [%.3f,%.3f]",
        small_sample_mean - t_95 * small_sample_sd / sqrt(small_sample_n),
        small_sample_mean + t_95 * small_sample_sd / sqrt(small_sample_n))
```
