---
title: "Oefening 6-04"
author: "Bert Van Vreckem"
date: "2020-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vcd)
library(ggmosaic)
```

For this exercise, we will use the dataset `survey` that is included in R. This dataset consists of a survey among students. The dataset can be imported as follows:

```{r}
library(MASS)
glimpse(survey)

# Make Exer and Smoke into ordinal variables, i.e. with ordering
survey$Exer <- factor(survey$Exer,
    levels = c("None", "Some", "Freq")
)
survey$Smoke <- factor(survey$Smoke,
    levels = c("Never", "Occas", "Regul", "Heavy")
)
```

You can view the entire dataset with `View(survey)` and get help on the contents with `?survey`.

> We want to investigate the relationship between some qualitative (nominal or ordinal) variables in this dataset. For each of the pairs listed below, follow these steps:
>
> - First think about what outcome you expect for the given combination of variables.
> - Compose a contingency table for the two variables. The (suspected) independent variable comes first.
> - Plot a graph of the data, e.g. a clustered bar chart, stacked bar chart of the relative frequencies, or a ``mosaic plot''
> - If you look at the graph, do you expect a large or a small value for the $\chi^2$ statistic? why?
> - Calculate the $\chi^2$ statistic and the critical value $g$ (for a significance level $\alpha = 0.05$)
> - Calculate the $p$-value
> - Do we have to accept or reject the null hypothesis? What does this imply for the relationship between the two variables?
>
> The variables to be investigated are listed below. The suspected independent variable is always mentioned first.
>
> The variables to be investigated are listed below.

| Independent variable    | Dependent variable                         |
|:------------------------|:-------------------------------------------|
| `Exer` (exercising)     | `Smoke` (smoking behavior)                 |
| `Sex` (gender)          | `Smoke`                                    |
| `W.Hnd` (dominant hand) | `Fold` (upper hand when folding your arms) |
| `Sex`                   | `W.Hnd`                                    |

# Exer vs Smoke

We stellen eerst de kruistabel op (eerst de afhankelijke, daarna de onafhankelijke variabele):

```{r}
exer_smoke <- table(survey$Smoke, survey$Exer)
addmargins(exer_smoke)
```

## Visualisation

Mosaic plot of the contingency table:

```{r}
ggplot(data = survey) +
    geom_mosaic(mapping = aes(x = product(Exer), fill = Smoke))
```

The plot indicates thtat the proportions of smoking behaviour between each category of variable `Sport` is similar. Consequently, we expect a relatively small value for $\chi^2$ and a weak association.

## Chi-squared test

The $\chi^2$-test goes as follows:

1. Hypotheses:

    - $H_0:$ There is no association between `Exer` and `Smoke`
    - $H_1:$ There is an association between `Exer` and `Smoke`

2. Calculate the test statistic:

    ```{r}
    smoke_exer_test <- chisq.test(
        survey$Smoke,
        survey$Exer
    )

    smoke_exer_test$statistic
    ```

3. Calculate the critical value:

    ```{r}
    alpha <- 0.05
    l <- (length(levels(survey$Smoke)) - 1) *
        (length(levels(survey$Exer)) - 1)

    g <- qchisq(p = 1 - alpha, df = l)
    g
    ```

4. Calculate the p-value

    ```{r}
    smoke_exer_test$p.value
    ```

5. Conclusion.

    ```{r}
    # Critical value $g$
    paste(ifelse(smoke_exer_test$statistic < g, "Accept", "Reject"), "the null hypothesis")
    # Probability value $p$
    paste(ifelse(smoke_exer_test$p.value > alpha, "Accept", "Reject"), "the null hypothesis")
    ```

    $\chi^2 < g$ and $p > \alpha$, so we can not reject the null hypothesis. Consequently, there is no significant association between the respondents' exercise habits (`Exer`) and smoking behaviour (`Smoke`).

Plot of this case (green: $\chi^2$, blue: $g$):

```{r}
xvals <- seq(0, 15, length = 200)
ggplot(mapping = aes(x = xvals)) +
    stat_function(fun = dchisq, args = list(df = l)) +
    geom_vline(xintercept = g, color = "blue") +
    annotate("text",
        x = g, y = .05,
        label = g, color = "blue"
    ) +
    geom_vline(xintercept = smoke_exer_test$statistic, color = "green") +
    annotate("text",
        x = smoke_exer_test$statistic, y = .02,
        label = smoke_exer_test$statistic, color = "green"
    )
```

## Cramér's V

```{r}
assocstats(exer_smoke)
```

Cramér's V is small ($V = 0,108$), which indicates a weak association. This corresponds with the result of the $\chi^2$-test.

# Sex vs Smoke

We first set up the contingency table (first the dependent, then the independent variable):

```{r}
gender_smoke <- table(survey$Smoke, survey$Sex)
addmargins(gender_smoke)
```

## Visualisation

Mosaic plot of the contingency table:

```{r}
ggplot(data = survey) +
    geom_mosaic(mapping = aes(x = product(Sex), fill = Smoke))
```

The graph indicates that the proportions of smoking behaviour within each category of variable `Sex` is similar. Consequently, we expect a relatively small value for $\chi^2$ and a weak association.

## Chi-squared test

The $\chi^2$-test goes as follows:

1. The hypotheses:

    - $H_0:$ There is no association between `Sex` and `Smoke`
    - $H_1:$ There is no association between `Sex` and `Smoke`

2. Calculate the test statistic:

    ```{r}
    gender_smoke_test <- chisq.test(
        survey$Smoke,
        survey$Sex
    )

    gender_smoke_test$statistic
    ```

3. Calculate the critical value:

    ```{r}
    alpha <- 0.05
    l <- (length(levels(survey$Smoke)) - 1) *
        (length(levels(survey$Sex)) - 1)

    g <- qchisq(p = 1 - alpha, df = l)
    g
    ```

4. Calculate the p-value

    ```{r}
    gender_smoke_test$p.value
    ```

5. Conclusion.

    ```{r}
    # Critical value $g$
    paste(ifelse(gender_smoke_test$statistic < g, "Accept", "Reject"), "the null hypothesis")
    # Probability value $p$
    paste(ifelse(gender_smoke_test$p.value > alpha, "Accept", "Reject"), "the null hypothesis")
    ```

    $\chi^2 < g$ and $p > \alpha$, so we can not reject the null hypothesis. Consequently, there is no significant association between the respondents’ gender (`Sex`) and smoking behaviour (`Smoke`).

Plot of this case (green: $\chi^2$, blue: $g$):

```{r}
xvals <- seq(0, 15, length = 200)
ggplot(mapping = aes(x = xvals)) +
    stat_function(fun = dchisq, args = list(df = l)) +
    geom_vline(xintercept = g, color = "blue") +
    annotate("text",
        x = g, y = .05,
        label = g, color = "blue"
    ) +
    geom_vline(xintercept = smoke_exer_test$statistic, color = "green") +
    annotate("text",
        x = smoke_exer_test$statistic, y = .02,
        label = smoke_exer_test$statistic, color = "green"
    )
```

## Cramér's V

```{r}
assocstats(exer_smoke)
```

Cramér's V is small ($V = 0,108$), which indicates a weak association. This corresponds with the result of the $\chi^2$-test.

# W.Hnd vs Fold

Deze oefening doen we iets compacter.

```{r}
whnd_fold <- table(survey$Fold, survey$W.Hnd)
chisq.test(whnd_fold)
assocstats(whnd_fold)
```

Again, we find no significant association.

# Sex vs W.Hnd

```{r}
gender_whnd <- table(survey$W.Hnd, survey$Sex)
chisq.test(gender_whnd)
assocstats(gender_whnd)
```

Again, we find no significant association.
