---
title: "Exercise 6-05 -- Aids"
author: "Bert Van Vreckem"
date: "2020-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggmosaic)
```

> Import the dataset `Aids2` from the package `MASS` that contains information on 2843 patients infected with AIDS in Australia before 1991. This dataset was discussed in detail by Ripley (2007).

```{r}
library(MASS)
glimpse(Aids2)
```

> Investigate whether there is an association between the variable (`Sex`) and the category of contamination (`T.categ`).
>
> - Proceed in the usual way: visualize the data, calculate $\chi^2$, $g$ and $p$-value ($\alpha = 0,05$), and finally formulate a conclusion.
> - Calculate the standardized residuals to determine which categories contain extreme values.

## Visualisation

```{r}
categ_sex <- table(Aids2$T.categ, Aids2$sex)
addmargins(categ_sex)

ggplot(data = Aids2) +
  geom_mosaic(mapping = aes(x = product(sex), fill = T.categ))
```

First and foremost, it is striking that there are far more men in the dataset than women. The transmission type is also different. For women, the mode is `blood` (contamination via blood transfusion, 37 observations), for men, it is `hs` (gay or bisexual contact, 2464 observations).

We can therefore expect that there will be an association between the two variables.

## Chi-squared test for independence

Testing procedure:

1. Formulate the hypotheses:
    - $H_0:$ There is no association between sex and transmission type
    - $H_1:$ There is an association between sex and transmission type
2. Determine significance level

    ```{r}
    alpha <- 0.05
    ```
    
3. Calculate the test statistic ($\chi^2$):

    ```{r}
    categ_sex_test <- chisq.test(categ_sex)
    categ_sex_test$statistic
    ```
4. Determine critical value:

    ```{r}
    g <- qchisq(1 - alpha, df = categ_sex_test$parameter[["df"]])
    g
    ```

5. Determine p-value:

    ```{r}
    categ_sex_test$p.value
    ```

6. Conclusion:

    - $g < \chi^2$, so it is well within the critical region
    - $p < \alpha$, which also indicates a significant association
    - Consequently, we can reject the null hypothesis. There is an association between gender and transmission type.

## Standardized residuals

```{r}
categ_sex_test$stdres
```

There are quite some values that can be considered to be extreme. Values smaller than -2 indicate underrepresented categories (e.g. `hs` for women), values greater than 2 overrepresented ones (e.g. `het` for women).

## Cochran's rule

As an extra, we look at whether Cochran's rule has been met. This is based on the expected values:

```{r}
categ_sex_test$expected
```

1. All expected values should be at least 1:

    ```{r}
    categ_sex_test$expected >= 1
    ```

    This rule is not met.

2. At least 80% of the expected values should be at least 5:

    ```{r}
    # How many values should be >= 5?
    ceiling(nrow(categ_sex) * ncol(categ_sex) * .80)
    categ_sex_test$expected >= 5
    ```
    
    Only 9 (instead of 13) values meet the condition, so the second rule is also not met.

If Cochran's rule is not met, this often has an impact on the 

Als de regel van Cochran niet voldaan is, heeft dat vaak een impact op de reliability of the results from the $\chi^2$-test. In this dataset, there are not enough observations about women.

## Referenties

Ripley, B. & Solomon, P. (2007). A note on Australian AIDS survival. Retrieved 2017-05-03, from
<https://pdfs.semanticscholar.org/7d23/36da875505e66ae983a271ee6cd83ce42677.pdf>
