---
title: "Exercise 6-11 en 12 -- Cats"
author: "Bert Van Vreckem"
date: "2020-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

>  Import data file “Cats.csv”.

```{r}
cats <- read_csv("../datasets/Cats.csv")
```

> 1. Perform a linear regression analysis on the variables body weight (`Bwt`, dependent variable) and heart weight (`Hwt`, independent variabele).
> 2. Draw a scatter plot of both variables.
> 3. Calculate and draw the regression line.
> 4. Calculate the correlation coefficient and the coefficient of determination.
> 5. Interpret the results from the previous steps.

## Data visualisation

The independent variable (`Hwt`) is mapped on the X-axis, the independent variable (`Bwt`) on the Y axis. We also draw the regression line:

```{r}
ggplot(data = cats, mapping = aes(x = Hwt, y = Bwt)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)
```

## Regression line

```{r}
#              dependent ~ independent
regr_cats <- lm(cats$Bwt ~ cats$Hwt)
```

The parameters of the regression line are:

```{r}
regr_cats
```

If we want to calculate this ourselves:

```{r}
mx <- mean(cats$Hwt)  # mean of x
my <- mean(cats$Bwt)  # mean of y
xx <- cats$Hwt - mx   # x - mx
yy <- cats$Bwt - my   # y - my
beta_1 <- sum(xx * yy) / sum(xx^2)
beta_0 <- my - beta_1 * mx
```

Which yields identical results.

## Correlation and determination

The correlation coefficient is:

```{r}
cor_Hwt_Bwt <- cor(x = cats$Hwt, y = cats$Bwt)
```

A correlation coefficient of about 0.804 indicates a strong linear and increasing relation.

**Please note!** You should not say a "strongly increasing relation". The correlation coefficient says nothing about the slope of the increase, only about how close the observations are to regression line.

The determination coefficient:

```{r}
cor_Hwt_Bwt^2
```

We conclude that about 64,66% of the variance in the observations can be explained by the regression line.

## Exercise 6.12 - Grouped by gender

First, an overview of the numbers of observations:

```{r}
addmargins(table(cats$Sex))
```

Compound plot wit regression lines:

```{r}
ggplot(data = cats, mapping = aes(x = Hwt, y = Bwt, color = Sex)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)
```

### Male

```{r}
m_cats <- cats[cats$Sex == 'M',]
regr_m_cats <- lm(m_cats$Bwt ~ m_cats$Hwt)
regr_m_cats
cor(x = m_cats$Hwt, y = m_cats$Bwt)
cor(x = m_cats$Hwt, y = m_cats$Bwt)^2
```

The correlation coefficient here is 0.79, indicating a strong linear and increasing relation. About 62.89% of the variance in the observations can be explained by the regression line.

### Female

```{r}
f_cats <- cats[cats$Sex == 'F',]
regr_f_cats <- lm(f_cats$Bwt ~ f_cats$Hwt)
regr_f_cats
cor(x = f_cats$Hwt, y = f_cats$Bwt)
cor(x = f_cats$Hwt, y = f_cats$Bwt)^2
```

The correlation coefficient here is 0,53, indicating a moderate linear and increasing relation. About 28,31% of the variance in the observations can be explained by the regression line.

There are far fewer observations of female cats than of males, so there is not such a clear pattern. Tomcats apparently also have a significantly heavier heart and body weight.

## Summary

```{r include=FALSE}
cats_b0 <- regr_cats$coefficients[["(Intercept)"]]
cats_b1 <- regr_cats$coefficients[["cats$Hwt"]]
mcats_b0 <- regr_m_cats$coefficients[["(Intercept)"]]
mcats_b1 <- regr_m_cats$coefficients[["m_cats$Hwt"]]
fcats_b0 <- regr_f_cats$coefficients[["(Intercept)"]]
fcats_b1 <- regr_f_cats$coefficients[["f_cats$Hwt"]]

cor_f <- cor(x = f_cats$Hwt, y = f_cats$Bwt)
cor_m <- cor(x = m_cats$Hwt, y = m_cats$Bwt)
```

To finish off, a summary of the most important results:

| Selectie     |    $\beta_0$ |    $\beta_1$ |             $R$ |             $R^2$ |
|:-------------|-------------:|-------------:|----------------:|------------------:|
| Hele dataset |  `r cats_b0` |  `r cats_b1` | `r cor_Hwt_Bwt` | `r cor_Hwt_Bwt^2` |
| Male         | `r mcats_b0` | `r mcats_b1` |       `r cor_m` |       `r cor_m^2` |
| Female       | `r fcats_b0` | `r fcats_b1` |       `r cor_f` |       `r cor_f^2` |
