---
title: "1 -- Getting started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculating

```{r}
1 + 2
3 * 4 + 5
3 * (4 - 5)
2^4
(1 + 7) / (3 + 8)
```

## Variables

To assign variables in R, we use `<-`.

```{r}
x <- 2
y <- 3
x + y
```

The `=` can also be used.
It is typically used when specifying parameters of a function.

Below you'll find 3 times the same command: We want rnorm() to generate 5 random values with a mean of 170 and a standard deviation of 15

```{r}
# 1st time: parameters in the correct order
rnorm(5, 170, 15)
# 2nd time: the names of the parameters are mentioned
rnorm(5, mean = 170, sd = 15)
# 3rd time: the order of the paramaters is no longer relevant
#           (because the names are mentioned)
rnorm(5, sd = 15, mean = 170)
```

The 2nd and 3rd notation is more readable.
Rstudio has autocompletion for the names of the parameters.

## Lists/vectors

Create lists ("c" stands for "collection")
```{r}
a <- c(1, 2, 3, 4) # assign a
b <- c(5, 6, 7, 8) # assign b
a; b # print a and b
```

You can also specify a *range* of values:
```{r}
c <- 0:5
c
```

Operations on a list: the operation is executed on every single element:
```{r}

a + 5 # every single element in list "a" is incremented with 5
a - 9
a * 4
a / 3
a ^ 2
```

Operations with 2 lists: executed one by one
```{r}
a + b
a - b
a * b
a / b
a ^ b
```

Selecting an element from a list:
_Remark_: the index starts at 1 (**not** at 0)!

```{r}
b[1]   # 1st element
b[2:3] # 2nd till 3rd element
```

## Functions for lists

```{r}
length(a) # number of elements, count
min(a)
max(a)
sum(a)
```

## Matrices

Below we create a 2 dimensional array or a *matrix*. You just give a list of values and the number of rows and columns.
_Remark_: the matrix is filled column by column (**not** row by row)!
```{r}
mat <- matrix(
      c(1, 2, 3,
        4, 5, 6), 
      nrow = 3,
      ncol = 2)
mat
```

Selecting from a matrix:

```{r}
mat[3,2]  # element on 3rd row, 2nd column
mat[,2]   # 2nd column
mat[1,]   # 1st row
mat[2:3,] # 2nd till 3rd row
```

## Data frames

Data frames typically consist of a series of columns, with the name of the column in the top row. Every column corresponds with a statistical variable. Variables can be qualitative (in R we call it a *factor*, *category* or *enumerated type*) or they can be quantitative.

We will use `mtcars`. It's a database that is available in R by default.

```{r}
?mtcars      # Shows the "man-page" for this dataset
View(mtcars)
nrow(mtcars) # number of rows
ncol(mtcars) # number of columns
```

Selecting elements from a frame:

```{r}
mtcars[1,2]  # element on 1st row, 2nd column
mtcars[3,]   # the entire 3rd row
mtcars[3:5,] # rows 3 till 5
mtcars[,2]   # 2nd column
mtcars[,2:4] # column 2 till 4
mtcars$cyl   # column with the name "cyl"
mtcars[c("mpg","hp")] # select only the columns 'mpg' and 'hp', returned as a table
```

You can also select a subset based on a kind of query. For example: "Return the number of cylinders (`cyl`) of the cars without an automatic gear (`am`)"

```{r}
mtcars$cyl[mtcars$am == 0]  
```

Or return all info of the subset with the syntax below. Remark the "comma" at the end!

```{r}
mtcars[mtcars$am == 0,]  
```

These "queries" can be combined with the logical operators *and* (`&`), *or* (`|`) and *not* (`!`):
```{r}
mtcars[mtcars$mpg >= 20 & mtcars$mpg <= 25,]
```
