---
title: "1 -- Getting Started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculating

```{r}
1 + 2
3 * 4 + 5
3 * (4 - 5)
2^4
(1 + 7) / (3 + 8)
```

## Variables

To assign variables in R, we use `<-`.

```{r}
x <- 2
y <- 3
x + y
```
The operator `=` can also be used.
It is typically used when specifying parameters of a function.

```{r}
# Provide parameters in the default order, without specifying the name
rnorm(50, 0, 1)
# The names of the parameters are used, the order of the parameters is no longer relevant
rnorm(50, mean = 0, sd = 1)
```

The second notation is more readable, and you don't have to remember the order of the parameters.
Rstudio has autocompletion for the names of the parameters.

## Lists / Vectors

```{r}
# Create 2 lists
a <- c(1, 2, 3, 4)
b <- c(5, 6, 7, 8)
# A list containing numbers 0 to 10 (range)
c <- 0:10

# Calculations using a list and a single number:
# Applies the operation to each element of the list
a + 5
a - 9
a * 4
a / 3
a ^ 2

# Calculations using two lists:
# Applies the operation using the corresponding elements from both lists
a + b
a - b
a * b
a / b
a ^ b
```

When selecting elements from a list, note that the index starts at 1!

```{r}
b[1]   # First element
b[2:3] # Second and third element
```

## List functions

```{r}
length(a) # Length
min(a)    # Smallest value
max(a)    # Biggest value
sum(a)    # Sum of all values
```

## Matrices

The following example creates a 2D array or *matrix*. 
You need to provide a list of all values, the number of rows and the number of columnbs. 
Note that the matrix column is filled in per column!

```{r}
mat <- matrix(
      c(1, 2, 3,
        4, 5, 6), 
      nrow = 3,
      ncol = 2)
mat
```

Selecting an element from a matrix:

```{r}
mat[3,2]  # Element from the third row, second column
mat[,2]   # Second kolom
mat[1,]   # First row
mat[2:3,] # Second and third row
```
