---
title: "5 -- Testing Procedures"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## The $Z$-test

The $Z$-test attempts to accept or reject a presumption regarding the *population mean* using a sufficiently large sample.

### The right-tailed $Z$-test

In general, we assume that superheroes rescue on average $\mu = 3.3$ people a day. Researchers want to verify this by taking a sample of 30 superheroes. For this sample, the mean is $\overline{x} = 3.483$. We can therefore suspect that on average, superheroes recue *more* people in a single day.

The standard deviation of the population is supposed to be known, $\sigma = 0.55$.

The testing procedure is as follows:

1. Formulate both hypotheses:
    - $H_0: \mu = 3.3$
    - $H_1: \mu > 3.3$
2. Determine the confidence level, e.g. $\alpha = 0.05$

    ```{r}
    # We have a sample with
    n <- 30      # sample size
    sm <- 3.483  # sample mean
    s <- 0.55    # standard deviation (supposed to be known)
    a <- 0.05    # significance level (selected by the researcher)
    m0 <- 3.3    # hypothetical population mean (H0)
    ```

3. Calculate the value of the test statistic, in this case $\overline{x} = 3.483$
4. Calculate the $p$-value and reject $H_0$ if $p < \alpha$.

    ```{r}
    p <- 1 - pnorm(sm, m0, s/sqrt(n))
    p
    if(p < a) {
      print("Reject H0")
    } else {
      print("Do not reject H0")
    }
    ```
    
    A different approach is to calculate the critical region, this is the area for which $H_0$ can be rejected. The bounday of this area is called the critical value $g$. Left of this value, you can not reject $H_0$ (region of acceptance), but you can reject the null hypothesis on the right of $g$ (critical region). The area of the region of acceptance is $1 - \alpha$, the area of the critical region is $\alpha$.
    
    ```{r}
    g <- m0 + qnorm(1-a) * s / sqrt(n)
    g
    # If the obtained sample mean is below g, we cannot reject H0
    if (sm < g) {
      print("Do not reject H0")
    } else {
      print("Reject H0")
    }
    ```

We can conclude that, based on $H_0$, the probability of obtaining a sample from the population with this sample mean is very small. We can therefore reject the null hypothesis for the selected significance level.    

Note that the conclusion should be identical for both methods (critical value and probability value)!

Below you can find a plot of this case:

```{r}
# x-values
x <- seq(m0-4*s/sqrt(n), m0+4*s/sqrt(n), length=200)
# x values left of g (region of acceptance)
i <- x <= g
# Density normal distribution
dist <- dnorm(x, m0, s/sqrt(n))

# Plot
ggplot() +
  # Gauss curve
  geom_line(aes(x, dist)) +
  # Plot the region of acceptance in green
  geom_polygon(aes(x = c(x[i],    g,                       g),
                   y = c(dist[i], dnorm(g, m0, s/sqrt(n)), 0)),
               fill = 'lightgreen',
               colour = 'black') +
  # Show the critical value
  geom_text(label = round(g, 3), aes(x = g, y = .5)) +
  geom_text(label = "Region of acceptance", aes(x = m0, y = .75)) +
  geom_text(label = "Critical region", aes(x = 3.6, y = .75)) +
  # Show the hypothetical population mean
  geom_vline(xintercept = m0) +
  geom_text(label = m0, aes(x = m0+.02, y = 2)) +
  # Show the sample mean
  geom_vline(xintercept = sm, colour = 'red') +
  geom_text(label = sm, aes(x = sm+.04, y = 3), colour = 'red') +
  theme(axis.title = element_blank())
```

### The left-tailed $Z$-test

Suppose that the result of the sample was $\overline{x} = 3.117$. In this case, we can suspect that the population mean is *smaller* than 3.3. We can verify this using a left-tailed $Z$-test.

1. Formulate both hypotheses:
    - $H_0: \mu = 3.3$
    - $H_1: \mu < 3.3$
2. Select significance level, e.g. $\alpha = 0.05$

    ```{r}
    # We have a sample with 
    n <- 30      # sample size
    sm <- 3.117  # sample mean
    s <- 0.55    # standard deviation (supposed to be known)
    a <- 0.05    # significance level (selected by researcher)
    m0 <- 3.3    # hypothetical population mean (H0)
    ```

3. Calculate the value of the test statistic, in this case $\overline{x} = 3.117$
4. Calculate the $p$-value and reject $H_0$ if $p < \alpha$.

    ```{r}
    p <- pnorm(sm, m0, s/sqrt(n))  # Note! NOT 1 - pnorm(...). Why?
    p
    if(p < a) {
      print("Reject H0")
    } else {
      print("Do not reject H0")
    }
    ```
    
    The critical value $g$ is now left of the mean. You cannot reject $H_0$ for a value on the right (region of acceptance), but you can for a value on the left of $g$ (critical region).
    
    ```{r}
    g <- m0 - qnorm(1-a) * s / sqrt(n)
    g
    # If the obtained sample mean is >g, we cannot reject H0
    if (sm < g) {
      print("Reject H0")
    } else {
      print("Do not reject H0")
    }
    ```

We can conclude that, based on $H_0$, the probability to obtain a sample with this sample mean is very small. We can therefore reject the null hypothesis for the selected significance level.

Plot of this case:

```{r}
# x-values for the plot
x <- seq(m0-4*s/sqrt(n), m0+4*s/sqrt(n), length=200)
# y-values (using the Gauss-curve)
dist <- dnorm(x, m0, s/sqrt(n))
# x values on the right of g
i <- x >= g

ggplot() +
  # Gauss curve
  geom_line(aes(x, dist)) +
  # Plot the region of acceptance in green
  geom_polygon(aes(c(x[i],    g),
                   c(dist[i], 0)),
               fill = 'lightgreen',
               colour = 'black') +
  # Show the critical value
  geom_text(label = round(g, 3), aes(x = g, y = .5)) +
  geom_text(label = "Region of acceptance", aes(x = m0, y = .75)) +
  geom_text(label = "Critical region", aes(x = 3, y = .75)) +
  # Show the hypothetical population mean
  geom_vline(xintercept = m0) +
  geom_text(label = m0, aes(x = m0+.02, y = 2)) +
  # Show the sample mean
  geom_vline(xintercept = sm, colour = 'red') +
  geom_text(label = sm, aes(x = sm+.04, y = 3), colour = 'red') +
  theme(axis.title = element_blank())
```

### The two-tailed $Z$-test

In this case, we do not want to test if the population mean is smaller/larger than a certain value, but instead we want to test if the obtained sample mean is similar to the population mean.

1. Formulate both hypotheses:
    - $H_0: \mu = 3.3$
    - $H_1: \mu \ne 3.3$
2. Select significance level, e.g. $\alpha = 0.05$

    ```{r}
    # We have a sample with
    n <- 30      # sample size
    sm <- 3.483  # sample mean
    s <- 0.55    # standard deviation (supposed to be known)
    a <- 0.05    # significance level (selected by researcher)    
    m0 <- 3.3    # hypothetical population mean (H0)
    ```

3. Calculate the value of the test statistic, in this case $\overline{x} = 3.483$
4. Calculate the probability value $p$ and reject $H_0$ if $p < \alpha/2$ (why do we divide by 2?).

    ```{r}
    p <- 1 - pnorm(sm, m0, s/sqrt(n))
    p
    if(p < a/2) {
      print("Reject H0")
    } else {
      print("Do not reject H0")
    }
    ```
    
    There are now two critical values:  $g_1$ is located on the left of the mean and $g_2$ on the right. The critical area still has a total area of $\alpha$.
    
    ```{r}
    g1 <- m0 - qnorm(1-a/2) * s / sqrt(n) # lower bound region of acceptance
    g2 <- m0 + qnorm(1-a/2) * s / sqrt(n) # upper bound region of acceptance
    c(g1, g2)
    
    # If the obtained sample mean is between g1 and g2, we cannot reject H0
    if (g1 < sm & sm < g2) {
      print("Do not reject H0")
    } else {
      print("Reject H0")
    }
    ```

Plot for this case:

```{r}
# Boundaries of the plot (x values)
x <- seq(m0-4*s/sqrt(n), m0+4*s/sqrt(n), length=200)
# y values using the Gauss curve
dist <- dnorm(x, m0, s/sqrt(n))
# x values between g1 and g2
i <- g1 <= x & x <= g2
# Probability of g1 and g2
dg1 <- dnorm(g1, m0, s/sqrt(n))
dg2 <- dnorm(g2, m0, s/sqrt(n))

# Plot
ggplot() +
  # Gauss curve
  geom_line(aes(x, dist)) +
  # Plot the region of acceptance in green
  geom_polygon(aes(c(g1, g1,  x[i],    g2,  g2),
                   c(0,  dg1, dist[i], dg2, 0)),
               fill = 'lightgreen',
               colour = 'black') +
  # Show the critical values
  geom_text(label = round(g1, 3), aes(x = g1, y = .3)) +
  geom_text(label = round(g2, 3), aes(x = g2, y = .3)) +
  
  geom_text(label = "Region of acceptance", aes(x = m0, y = .75)) +
  geom_text(label = "Critical region", aes(x = 3, y = .75)) +
  geom_text(label = "Critical region", aes(x = 3.6, y = .75)) +
  
  # Show the hypothetical population mean
  geom_vline(xintercept = m0) +
  geom_text(label = m0, aes(x = m0+.02, y = 2)) +
  # Show the sample mean
  geom_vline(xintercept = sm, colour = 'red') +
  geom_text(label = sm, aes(x = sm+.04, y = 3), colour = 'red') +
  theme(axis.title = element_blank())
```

If we do not say a priori whether the actual population mean will be smaller or greater, the obtained sample mean could be obtained from this population. In other words: we can *not* reject the null hypothesis.
 

## The $t$-test

If the conditions applying the $z$-test do *not* hold (e.g. sample size is too small and/or standard deviation of the population is unknown), the $t$-test could be used instead. Similar to the $Z$-test, there are three alternatives: left-tailed, right-tailed or two-tailed. In this section, we will only illustrate the right-tailed test.


### Right-tailed $t$-test

Suppose that the researchers did not have enough time to take a sufficiently large sample, and only have $n = 20$ observations. The obtained sample mean is the same, 
$\overline{x} = 3.483$. The standard deviation of the sample is $s = 0.55$.

Can we - given these conditions - conclude the same for a same significance level of $\alpha = 0.05$, namely that superheroes rescue *more* than 3.3 people a day?

The testing procedure is as follows:

1. Formulate both hypotheses:
    - $H_0: \mu = 3.3$
    - $H_1: \mu > 3.3$
2. Select significance level, e.g. $\alpha = 0.05$

    ```{r}
    # We have a sample with
    n <- 20      # sample size
    sm <- 3.483  # sample mean
    ss <- 0.55   # standard deviation sample
    a <- 0.05    # significance level (selected by researcher)
    m0 <- 3.3    # hypothetical population mean (H0)
    ```

3. Calculate the value of the test statistic, in this case $\overline{x} = 3.483$
4. Calculate the probability value $p$ and reject $H_0$ if $p < \alpha$. Note: the function `pt` does not allow for using a custom mean and standard deviation, therefore we have to normalize the values ourselves.

    ```{r}
    p <- 1 - pt((sm - m0) / (s/sqrt(n)), df = n-1)
    p
    if(p < a) {
      print("Reject H0")
    } else {
      print("Do not reject H0")
    }
    ```
    
    For calculating the critical value, we use the `qt` function with $n - 1$ degrees of freedom.
    
    ```{r}
    g <- m0 + qt(1-a, df = n-1) * s / sqrt(n)
    g
    # If the obtained sample mean is below g, we cannot reject H0
    if (sm < g) {
      print("Do not reject H0")
    } else {
      print("Reject H0")
    }
    ```
    
Plot for this case:

```{r}
# Boundaries of plot (x values)
x <- seq(m0-4*ss/sqrt(n), m0+4*ss/sqrt(n), length=200)
# y-values according to gauss curve of the t-distribution
dist <- dt((x-m0)/(ss/sqrt(n)), df = n-1) * ss/sqrt(n)
# x values left of g
i <- x < g   
# value of probability function for g
dg <- dt((g-m0)/(ss/sqrt(n)),df=n-1) * ss/sqrt(n)

# Plot
ggplot() +
  # Gauss curve
  geom_line(aes(x, dist)) +
  # Plot region of acceptance in green
  geom_polygon(aes(x = c(x[i],    g, g),
                   y = c(dist[i], dg, 0)),
               fill = 'lightgreen',
               colour = 'black') +
  # Show critical value
  geom_text(label = round(g, 3), aes(x = g, y = .005)) +
  geom_text(label = "Region of acceptance", aes(x = m0, y = .0075)) +
  geom_text(label = "Critical region", aes(x = 3.7, y = .0075)) +
  # Show the hypothetical population mean
  geom_vline(xintercept = m0) +
  geom_text(label = m0, aes(x = m0+.02, y = .02)) +
  # Show the sample mean
  geom_vline(xintercept = sm, colour = 'red') +
  geom_text(label = sm, aes(x = sm+.04, y = .03), colour = 'red') +
  theme(axis.title = element_blank())
```

In other words, although we get similar results in our sample, we cannot draw the same conclusion. Because our sample is too small, there is a large uncertainty whether the value of the sample mean is extreme enough to reject the null hypothesis.

### Use of the function `t.test`

In R, there is also a function `t.test` that can be used to apply the test on a collection of observations.

```{r}
test_sample <- c(                # sample
  3, 2, 3, 1, 10, 4, 2, 7, 3, 0,
  3, 1, 2, 3,  4, 0, 3, 8, 3, 7)
a <- 0.05                        # significance level
m0 <- 3.3                        # hypothetical population mean (H0)

t.test(test_sample, 
       alternative = "greater",  # right-tailed test
       mu = m0,                  # hypothesis about mean
       conf.level = 1 - a)       # confidence level
```

In this case, we conclude that the sample mean of 3.45 does not indicate that we can reject the null hypothesis, as the probability value $p = 0.4013$ is larger than the significance level $\alpha = 0.05$.
