---
title: "Exercise 7.2 - Budget"
author: "Bert Van Vreckem"
date: '2020-05-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(TTR)
library(forecast)
```

> The file `Budget.csv` contains the turnover, advertising budget and GNP (Dutch: BNP) of a medium sized compay per quarter between 1981 and 2005.

```{r}
budget <- read_csv("datasets/Budget.csv") %>%
  mutate(Date = parse_date(Quarter, format = "%b-%y"))

glimpse(budget)
```

In the dataset, the column Quarter is formatted as "Mar-81", i.e. the abbreviated name of the month, and the year (without the century). We convert these into a valid date with `mutate()`.

Let's take a look at the time series `Revenue`:

```{r}
ggplot(data = budget, mapping = aes(x = Date, y = Revenue)) +
  geom_line()
```

For several of the calculations in this exercise, it is useful to create a time series object (`ts`). We do this for the Revenue first:

- `frequency` indicates that the time series has four time units per cycle (a year has 4 quarters)
- `start` denotes the time series starts in the first tume unit of cycle 1981
- `end` denotes the last observation is for the fourth time unit of cycle 2005

```{r}
revenue <- ts(budget$Revenue, frequency = 4,
              start = c(1981, 1), end = c(2005, 4))
revenue
```

R has apparently deduced that the dataset contains quarterly figures!

In what follows, we'll investigate the variable `Revenue` more closely. You can repeat the exercise for yourself with `AdBudget` and `GNP`.

# Moving average

> Calculate the simple moving average for periods of 4 and 12 months, respectively. Use the function `SMA()`. Plot a line chart of the observations, SMA(4) and SMA(12).

```{r warning=FALSE}
budget_sma <- budget %>%
  mutate(revenue_sma_4 = SMA(Revenue, n = 4),
         revenue_sma_12 = SMA(Revenue, n = 12))
ggplot(data = budget_sma, mapping = aes(x = Date, y = Revenue)) +
  geom_line() +
  geom_line(mapping = aes(y = revenue_sma_4), color = "orange") +
  geom_line(mapping = aes(y = revenue_sma_12), color = "blue")
```

# Regression

> Which technique we have discussed earlier in the course is also suitable to
make predictions about future values of X? Work this out using the appropriate function and plot the result in the graph.

We are talking about regression here. A time series that continues to fluctuate around a certain "equilibrium point" can be approximated with a regression law.

```{r}
ggplot(data = budget, mapping = aes(x = Date, y = Revenue)) +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ x)
```

The parameters of the regression line can be calculated with `lm()`:

```{r}
lm(revenue ~ time(revenue))
```

Remark that we don't use the data in the tibble (variable `budget`, created with the function `read_csv()`), but the time series object `revenue`. The function `lm()` expects a formula in the format `y ~ x`. `y` is the time series, `x` is the time on the x-axis. The date in the `budget` tibble can't be used for this. One could create a time index (`1:n`) instead, but the function `time()` always gives a useable result for `lm()` when provided with a time series.

# Forecasting

>  Use function `forecast()` to predict the next 10 periods. Draw these on the graph.

```{r}
revenue_fc <- forecast(revenue, h = 10)
revenue_fc
plot(revenue_fc)
```

This plot was not generated using `ggplot()`, but with the `plot()` function from base-R. `ggplot2` was designed specifically to work with tibbles and data frames and cannot be used with time series objects. For plotting time series, we often need to revert to the base-R plotting system. You could convert the time series data into a tibble, but that would lead us too far.

> Is one of these techniques useful to make a forecast for this time series?

The time series seems to fluctuate around about 950, but there is a lot of variance between individual observations. There also seems to be a recurring seasonal trend in the time series, which makes linear regression less suitable for predictions.

# Seasonal fluctuations

> Use the `decompose()` function to, well, decompose the time series and to get an idea of any long term trend and seasonal fluctuations.

The result of `decompose()` is:

```{r}
plot(decompose(revenue, type = "multiplicative"))
```

The seasonal trend estimate is a vector of as many numbers as the specified `frequency`, so here 4. If we had passed parameter `type = "additive"`, then the estimate of the seasonal trend is added to the estimate on basis of the underlying linear trend. With type "multiplicative" like here, the estimate is multiplied.

# Exponential smoothing

> Calculate the exponential moving average (EMA) using the `HoltWinters()` method. Again, use `forecast()` to make a prediction for 20 quarters. Use start values $s_1 = x_1$ and $\alpha$ the value calculated by R. Plot the result along with the observed time series.

```{r}
revenue_ema <- HoltWinters(revenue, beta = FALSE, gamma = FALSE,
                           seasonal = "multiplicative", s.start = revenue[1])
plot(revenue_ema)
revenue_ema_fc <- forecast(revenue_ema, h = 20)
plot(revenue_ema_fc)
```

Simple exponential smoothing works from the assumption that the time series fluctuates around a constant, and consequently has no trend (neither linear, nor seasonal).

\[x_t = a + u_t\]

with $a$ a constant and $u_t$ noise. The prediction, therefore, is also a constant:

```{r}
revenue_ema_fc$mean
```

> Now do the same with $\alpha = 0,1$

```{r}
revenue_ema_2 <- HoltWinters(revenue, alpha = 0.1, beta = FALSE, gamma = FALSE,
                           seasonal = "multiplicative", s.start = revenue[1])
plot(revenue_ema_2)
revenue_ema_fc_2 <- forecast(revenue_ema_2, h = 20)
plot(revenue_ema_fc_2)
```

The prediction becomes:

```{r}
revenue_ema_fc_2$mean
```

# Double exponential smoothing

> Do the same with double exponential smoothing. Bootstrap with $s_1 = x_1$ and $b_1 = \frac{x_n−x_1}{n-1}$, $\alpha = 0,05$ and $\beta = 0,2$. Plot the result.

Double exponential smoothing assumes that the data has a linear trend and can be modelled with equation:

\[x_t = a + bt + u_t\]

with met $a$ again a constant (intercept with the y-axis), $b$ the slope and $u_t$ noise.

```{r}
n <- length(revenue)
revenue_dema <- HoltWinters(revenue, alpha = 0.05, beta = 0.2, gamma = FALSE,
                            seasonal = "multiplicative",
                            s.start = revenue[1], 
                            b.start = (revenue[100]-revenue[1])/100)
plot(revenue_dema)
```

> Use double exponential smoothing to make a prediction for 20 quarters. Plot the values. Is this technique better or worse than the previous one for this dataset?

```{r}
revenue_dema_fc <- forecast(revenue_dema, h = 20)
plot(revenue_dema_fc)
```

The estimate of the slope $b$ is adapted each time unit. Because there is so much fluctuation in the data, $b$ will alternate between positive and negative values, depending on where we are in the seasonal cycle. The forecast will therefore be a rising or falling line, which makes little sense for this data.

> Play around with the values for $\alpha$ and $\beta$ and look at the results, both for simple and double exponential smoothing.

Do this yourself! Use the code above and adapt. The result will be that the predictions are higher or lower with single EMA, or that we will see a rising of falling line with a different slope with double EMA.

# Seisonal trend

Holt-Winters' method, or triple exponential smoothing, assumes the observations follow this model:

\[x_t = (a + bt) s_t + u_t \]

with $a$, $b$ and $u_t$ as before and $s$ a list of numbers as long as the `frequency`, $k$ (= 4 in this time series). At time $t$, the element $t \mod k$ from this list, $s_t$, is multiplied with the original value.

> Use the `HoltWinters()` function without a trend, i.e. set $\beta =$ `FALSE`. Bootstrap with $\alpha = 0.05$ and $\gamma = 0.9$. Plot the result.

```{r}
revenue_hw <- HoltWinters(revenue, alpha = 0.05, beta = FALSE, gamma = 0.9,
                          seasonal = "multiplicative")
plot(revenue_hw)
```

We now see that the estimates are pretty close to the observed values, at least compared with the previoous models.

The estimates of the parameters can be retrieved with:

```{r}
revenue_hw
```

We see an estimate for the underlying constant, `a` = `r revenue_hw$coefficients[["a"]]` and seasonal parameters (`s1` to `s4`). The first quarter always has the highest revenue, the third quarter the lowest. The second and fourth quarter are "normal", i.e. close to `a`.

> Recalculate the forecast for 20 time units. Plot the results. Is this technique better or worse than the previous ones for this dataset?

```{r}
revenue_hw_fc <- forecast(revenue_hw, h = 20)
plot(revenue_hw_fc)
```

The predicted values are much closer to the observations. You can see the seasonal trend in the forecasted values. The forecasts are calculated from `a` and `s1` to `s4`. For the first quarter of a year, that's `a * s1`, for the second `a * s2`, etc.

> Play around with the values for $\alpha$ and $\beta$ and look at the results.

Do this yourself! Adapt the values and look at the results on the chart.

> Use `HoltWinters()` with the smoothing parameters $\alpha$ and $\gamma$ as determined by R. Turn off long term trend, i.e. set `beta = FALSE`. Plot the result.

```{r}
revenue_hw_2 <- HoltWinters(revenue, beta = FALSE, seasonal = "multiplicative")
plot(revenue_hw_2)
# Model, berekende parameters
revenue_hw_2
```
The smoothing parameter $\alpha$ calculated by R, and the estimates for seasonal parameters `s1` to `s4` are close to the values in the previous model. Parameter $\gamma$ is close to 0, while we had set it to 0.9 previously. Apart from that, the model seems to be performing similarly. 

```{r}
revenue_hw_2_pr <- predict(revenue_hw_2, n = 20)
plot(revenue_hw_2_pr)
```

We see forecasted values similar to the ones in the previous model.
