---
title: "Exercise 5.3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

> It is being said that introducing a "binding recommendation on continuation of studies" (refusing enrollment in the next academic year if a student did not complete a certain level of credits) has a positive effect on the study efficiency and success rate. Before the introduction of binding recommendations, the number of completed credits per student per year was 44 with a standard deviation of 6.2. After the introduction, a sample of 72 random students has an average number of completed credits of 46.2.
> Test whether there is evidence that the introduction of binding recommendations has improved the success rate among students. Calculate the critical value for a significance level of $\alpha = 2.5\%$.


## Data

```{r}
m <- 44     # Population mean
s <- 6.2    # Standard deviation population
a <- 0.025  # Significance level

n <- 72     # Sample size
sm <- 46.2  # Sample mean
```

## Plot for this casus

```{r}
sn <- s/sqrt(n)
ggplot(data = data.frame(x = c(m-4*sn, m+4*sn)), mapping = aes(x)) +
  stat_function(fun = dnorm, n = 201, args = list(mean = m, sd = sn)) +
  geom_vline(xintercept = m) +
  geom_vline(xintercept = sm, color = "orange")
```

## Hypotheses

The sample is sufficiently large (n > 30) and $\sigma$ is known. We will apply a right-tailed z-test. The hypotheses are:

\[ H_0: \mu = 44 \]
\[ H_1: \mu > 44 \]

# Critical value

> Calculate the critical value.

```{r}
g <- m + qnorm(1-a) * s/sqrt(n)
g

sn <- s/sqrt(n)
ggplot(data = data.frame(x = c(m-4*sn, m+4*sn)), mapping = aes(x)) +
  stat_function(fun = dnorm, n = 201, args = list(mean = m, sd = sn)) +
  geom_vline(xintercept = m) +
  geom_vline(xintercept = sm, color = "orange") +
  geom_vline(xintercept = g, color = "red")
```

The sample mean $\overline{x}$ is inside the critical region ($g < \overline{x}$), so we can reject the null hypothesis.

# Probability value

> Do the same by calculating the $p$-value.

```{r}
p <- 1-pnorm(sm, m, s/sqrt(n))
p
```

The Probability value is smaller than the significance level ($p < \alpha$) so we can reject the null hypothesis.

# Interpretation of alpha

>  Explain what $\alpha = 2.5 \%$ means.

The significance level denotes the Probability of making a Type I error if we assume that the null hypothesis is true. In other words, this is the probability, if $H_0$ is true, to have a sample with such extreme values that the calculated sample mean is inside the critical region.

