---
title: "Exercise 1.2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this exercise we work with the built-in data frame `mtcars`.
    
1. Use built-in R functions to retrieve information about the dataset

    ```{r}
    #View(mtcars)  # Contents 
    #?mtcars       # More information
    ```

2. Return the value of the first row, second column

    ```{r}
    mtcars[1,2]
    ```

3. Return the number of rows and the number of columns

    ```{r}
    nrow(mtcars) # Number of rows
    ncol(mtcars) # Number of columns
    ```

4. Return only the column containing the definitions of the cylinders

    ```{r}
    # The whole column
    mtcars$cyl
    # number of cylinds for cars without automatic transmission
    mtcars$cyl[mtcars$am == 0]  
    ```

To obtain a data frame with the two columns `mpg` and `hp`, add the column names to an index vector with a single square bracket operator.

    ```{r}
    mtcars[c("mpg","hp")]
    ```

Try to find out how you can determine a row record of the built-in dataset `mtcars`.

    ```{r}
    mtcars[3,]               # Third row
    mtcars["Datsun 710",]    # Row with provided name
    mtcars[mtcars$cyl == 6,] # Rows compliant to provided condition
    ```
