---
title: "Exercise 1.6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Remark:** Before running this code in RStudio, first do (from the menu): *Session > Set Working Directory > To Source File Location*. This allows loading the .csv file from the current directory.

Import the file \texttt{rainforest.csv} in R.

For this data file, try counting how many rows there are for each species that are complete (i.e. do not contain n.a.). 
You can use `with`, `table`, and `complete.cases` to achieve this.


1. Import the file

    ```{r}
    # Load data from csv
    rainforest <- read.csv("../datasets/rainforest.csv", sep = ",")
    # Allows the columns to be accessed directly instead of with table$column notation
    attach(rainforest)

    #View(rainforest) # View contents
    ```
    
2. Select complete records 

    ```{r}
    # Array indicating for each record if it is complete (TRUE/FALSE)
    ok <- complete.cases(rainforest)
    # Selection from rainforest containing only complete records.
    # Only the column "species" is stored.
    complete <- rainforest[ok,c("species")]
    ```
    
3. Create a frequency table containing the number of complete cases for each species

    ```{r}
    table(complete)
    ```
  
**Conclusion:** Only *B. myrtifolia* has four complete cases, the other species don't have any.
