\chapter{The Research Process}
\label{ch:research_process}

\section{Learning Goals}
\label{sec:onderzoeksproces-leerdoelen}

By the end of this chapter you must be able to:

\begin{itemize}
    \item Define the terms in this chapter;
    \begin{itemize}
        \item scientific method and empirical validation
        \item fundamental and applied research
        \item variable, value
        \item measurement levels: nominal, ordinal, interval, ratio
        \item population, sample, sampling frame
        \item random vs.~ select sample
        \item stratified sample
        \item sampling errors: accidental/systematic, sampling/non-sampling error
    \end{itemize}
    \item Based on the description of a case or situation, determine whether or not they meet the characteristics of the scientific method or empirical validation;
    \item Identify and explain scientific research objectives;
    \item Identify and explain the different steps in a research process;
    \item List the four measurement levels, formulate the characteristics of each and give an example;
    \item Determine the measurement level for a given variable;
    \item List the types of sampling errors, formulate the characteristics of each and give an example;
    \item Based on the description of a sampling method:
    \begin{itemize}
        \item make a distinction between a random and a selective sampling method;
        \item identify and explain the type of sampling error(s) made;
        \item make proposals to improve the sampling method;
    \end{itemize}
\end{itemize}

\section{The Scientific Method}
\label{sec:onderzoeksproces-wetenschappelijke-methode}

There are several ways to gather knowledge:

\begin{enumerate}
    \item The scientific method    
    \item The non-scientific method
\end{enumerate}

\paragraph{Non-Scientific} Several versions of non-scientific reasoning exist:
\begin{description}
	\item [Authoritarian] Someone is considered to be an authority on a certain subject and to be trustworthy. Every claim this person makes is regarded as the truth.
	\item [Deductive] Given a set of assumptions, logical reasoning leads to certain conclusions. This way, correct results \emph{may} be found, but these depend solely on the truth of the assumptions. These are not researched empirically, though.
\end{description}

\paragraph{Scientific} 
A characteristic of the \textsl{scientific method} is \textbf{empirical validation}: based on experiments and direct observation. A claim is only valid if it corresponds to what is observed.

\begin{exercise}
Try to prove that pigs can fly, starting from both non-scientific and scientific reasoning.
\end{exercise}

Empirical research may have several goals:
\begin{enumerate}
    \item Exploration: Does something exist, or does a certain event occur?
    \item Description: What are the properties of this event?
    \item Prediction: Is a certain event related to another one and is it possible to predict it?
    \item Verification: Is it possible to completely predict a certain event based on other observations?
\end{enumerate}

\paragraph{Research Objectives}

Most research objectives can be categorized into:

\begin{description}
  \item[Generalization] Often research is conducted on a limited \emph{sample} of the group that is being studied, the \emph{population}. If conclusions for this sample also hold for the population, we have found a valid generalization.
  \item[Specialization] Applying general knowledge to specific instances or domains. A lot of applied research can be classified under this category.
\end{description}

There are two types of generalization:

\begin{enumerate}
	\item About a single phenomenon;
	\item About relations between phenomena.
\end{enumerate}

There are three reasons why relationships between phenomena are so important:

\begin{enumerate}
	\item A complete understanding of the phenomenon;
	\item Relations may enable prediction;
	\item Revealing causal relations.
\end{enumerate}

\paragraph {Fundamental vs. Applied Research}
Depending on the research objective, we speak of either fundamental or applied research.

\emph{Basic or Fundamental Research} is typically conducted at universities. Researchers try to expand the existing knowledge in their field. In computer science, for example, it can be about developing new algorithms. Fundamental research does not primarily take practical applications into account. You could say that fundamental research focuses primarily on developing new solution methods, and only afterwards on what problems can be solved more efficiently by using these methods. It is difficult to predict what impact (in particular added financial value) fundamental research results will have. In the best case it could change the world, in the worst case there is no practical application.

\emph{Applied research} starts with a concrete problem, typically in a business context. Researchers must first familiarize themselves with the specific problem domain. Then they can look for the most suitable method to solve that problem. They must therefore also be aware of the state-of-the-art within the relevant fundamental research. The added value of applied research is usually easier to measure, but the impact is limited to the company or organization for which the research was conducted.

\section{Basic Concepts in Research}
\label{sec:research-process-basic-concepts}

\paragraph{Measurement Levels}

In quantitative research, we try to comprehend a phenomenon by measuring it.

\begin{definition}[Variable]
     A general property of an object that allows us to distinguish it (e.g.~length, mass, etc.).
\end{definition}

\begin{definition}[Value]
    The specific property of a specific object. The measurement assigned to a variable (e.g.~1.83m, 78kg, etc.).
\end{definition}

In statistical analysis, a classification of variable types is used that helps determining which statistical methods can be applied to the variable in question. 
This subsection introduces a common taxonomy for the possible measurement levels of a variable.

\subparagraph{Qualitative Scales}

Qualitative values are not necessarily numeric.
They always have a limited number of possible values.

\begin{description}
    \item [Nominal] The variable can only take a limited number of values, and there is no logical order between them. Some examples include gender, nationality, biological species, etc.
    \item [Ordinal] The variable can take a limited number of values, but there is a logical ordering between them. E.g.~Likert scale when measuring opinion (from \emph{strongly disagree} to \emph{strongly agree}), military rank, etc.	
\end{description}

\subparagraph{Quantitative Variables}

Quantitative variables often contain the result of a measurement.
They are usually expressed using a number and unit of measurement.
Quantitative variables can have many values, and often values are unique.

\begin{description}
    \item [Interval] A numerical variable that allows for measuring the difference between items, but not the ratio between them. There is no meaningful zero value. E.g.~temperature.
    \item [Ratio] A numerical variable with a non-arbitrary zero point, that allows for meaningful ratios between values (e.g.~``twice as long,'' ``5\% higher,'' etc.). E.g.~mass, age, etc.
\end{description}

\begin{exercise}
    Think of some other examples for all 4 measurement levels.
\end{exercise}

\paragraph{The Research Process}

The research process can largely be subdivided into six main phases:

\begin{enumerate}
	\item Formulating the problem statement: what is the research question?
	\item Defining the exact information need: what are the specific questions we need to ask, what do we need to measure?
	\item Conducting the research: surveys, simulations, experiments, \dots
	\item Processing result data, nowadays typically using statistical software
	\item Analyzing the results: applying statistical methods
	\item Writing conclusions and reporting on the conducted research
\end{enumerate}

\begin{definition}[Causal Relation]
    Two variables have a \emph{causal relation} when a change in the first variable reliably causes an associated change into the second one, provided that all other potential causes are eliminated. The first variable is called the \emph{independent variable}, the second the \emph{dependent variable}.
\end{definition}

A relationship is not always readily apparent, and sometimes we must look further than the actual values of variables before drawing conclusions.

\begin{example}
    Inspect the values of Table~\ref{tab:pepsi-coca}, which are the results of a taste test. 
    A hundred persons were asked to blindly taste a kind of cola and indicate whether they liked it or not.
    
    Initially, one could think that Pepsi is better because more people liked it. 
    However, that would not be correct. We need to look at the relative numbers. 
    Out of $70$ people that tasted Pepsi, $56$ found it tasty, or $80\%$ ($\frac{56}{70} = 0.8$). 
    Out of the $30$ that tasted Coca cola, $24$ found it tasty, which is also $80\%$ ($\frac{24}{30} = 0.8$). 
    Consequently, there is no difference between the results for Pepsi vs.~Coca Cola.
\end{example}    

\begin{table}
    \centering
    \begin{tabular}{l|cc|c}
                  & Pepsi & Coca Cola & Total \\
        \midrule
        Like      &  56   &    24     &   80  \\
        Dislike   &  14   &     6     &   20  \\
        \midrule
        Total     &  70   &    30     &  100
    \end{tabular}
    \caption[Results of a taste test between Pepsi and Coca Cola.]{Results of a taste test between Pepsi and Coca Cola. A hundred persons were asked to blindly taste a kind of cola (Pepsi or Coca Cola) and to say whether they liked it or not.}
    \label{tab:pepsi-coca}
\end{table}

\section{Sample Testing}
\label{sec:onderzoeksproces-steekproefonderzoek}

A reason for conducting quantitative research is being able to make statements that give a representative picture of the reality. A sample is often used for this. A sample is a subset of the full population for the purpose of measuring certain characteristics of that population.

\subsection{Sample and Population}

\begin{definition}[Population]
    A \index{population}\emph{population} is the collection of \textbf{all} elements (objects/people/\ldots) that you want to investigate.
\end{definition}

\begin{definition}[Sampling Frame]
    A \index{sampling frame}\emph{sampling frame} is a list that includes all members of a population to be investigated.
\end{definition}

\begin{definition}[Sample]
    A \index{sample}\emph{sample} is a subset of the population on which the researchers will effectively take measurements, or on which they will collect specific information.
\end{definition}

\begin{figure}
    \begin{center}
        \begin{tikzpicture}[scale=.55]
        \fill[hgyellow] (2,2) ellipse (4cm and 2cm) ;
        \fill[hgorange] (1.5,2) ellipse (2cm and 1cm) ;
        \node[draw=none,minimum size=1cm,inner sep=0pt] at (3,0.5) {population};
        \node[draw=none,minimum size=1cm,inner sep=0pt] at (2.5,2) {sample};
        \end{tikzpicture}
    \end{center}
    \caption{Sample and Population}
    \label{img:populatie-steekproef}
\end{figure}

There are a number of reasons for taking a sample:

\begin{itemize}
    \item Population is too large to collect the required information from the entire population.
    \item Performing a measurement is too expensive, which would cause the experiment to become too costly.
    \item When time is limited, it is often faster to investigate a subgroup.
    \item Taking a sample is in any case easier than examining the entire population.
    \item \dots
\end{itemize}

After taking the necessary measurements, or collecting the necessary information, the researchers will draw a conclusion. Under certain strict conditions, the results for the sample may also be generalized to the entire population. In other words, if you have taken the sample correctly, you can assume that what you observe within the sample will apply to the entire population. At a later point we will go into more detail about what these conditions are.

\subsection{Choosing a Sampling Method}

\subsubsection{Random Sample}

There are different techniques for taking a sample from a population. In an ideal world, the best way to set up a sample is as follows:

\begin{enumerate}
    \item \textbf{Define the Population}: what is the precise target group? This is closely related to the research problem. This is a very important step that you should not go over lightly. It is important to describe the target group / population as good as possible. Elements of interest are, for example, social, demographic or physical characteristics such as gender, age, place of residence, \dots
    \item \textbf{Determine the Sampling Frame}: if the target group is well defined, you can compile a list of all elements that are part of the population. This list could consist of people, companies, products, etc., depending on the research question.
    \item \textbf{Random sample}: the researchers then use the sampling frame to select a number of elements \textit{at random} from which they will collect information.
\end{enumerate}

It is extremely important that the selection process is effectively random, that is, every element of the population has an equal chance of being selected. We will investigate the reason for this later in this course, when we talk about the Central Limit Theorem (cfr. Section~\ref{sec:central-limit-theorem}).

\begin{definition}[random sample]
    We call a sample \index{sample!random}\emph{random} if every element has an equal chance of being selected.
\end{definition}

Unfortunately, it is often impractical to take a random sample. Sometimes it is practically impossible to set up a sampling frame, even if the target group is well defined.

\begin{example}
    Every year, the imec research center in Flanders conducts research regarding the use of media and technology. The population is defined as all Flemish people older than 16. 
    
    On January 1st 2019, the Flemish Region had approximately 6.6 million inhabitants, excluding Dutch speakers in Brussels~\autocite{Statbel2019}. You can imagine that as a research institution it is not possible to obtain a list with the names and contact details of all those residents. The government might be the only one able to compile such a list, but the law on the protection of privacy does not allow it to be shared with other organizations.
    
    For this reason, imec is forced to use a different method for taking a sample.
\end{example}

\subsubsection{Stratified Samples}

Sometimes the population is very diverse on a number of important characteristics. Therefore, the population as a whole is classified into a number of non-overlapping and homogeneous strata or classes, e.g. age category, gender, level of education, etc.

\begin{definition}[Stratified Sample]
    \index{sample!stratified}A \textbf{stratified} sample is a sample in which the proportion of each subpopulation in the sample is equal to the proportion of the subpopulation in the full population.
\end{definition}

\begin{example}
    Table~\ref{tab:frequenties-populatie} illustrates the absolute number of men and women for different age groups of a population.
    We cannot question all members of the population, but if we take a sample where the men/women and age category are relatively equivalent to the population, we have taken a stratified sample (cfr. Table~\ref{tab:frequenties-steekproef}).
\end{example}

\begin{table}
    \centering
    \begin{tabular}{l|cccc|c}
        & \multicolumn{4}{c|}{\textbf{Age}} & \\
        Gender & $\le 18$ & $]18,25]$ & $]25, 40]$ & $> 40$ & Total\\
        \hline
        Woman & 500 & 1500 & 1000 & 250 & 3250 \\
        Man   & 400 & 1200 & 800 & 160 & 2560\\
        \hline
        Total & 900 & 2700 & 1800 & 410 & 5810
    \end{tabular}
    \caption{Number of men and women for different groups within a fictional population.}
    \label{tab:frequenties-populatie}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{l|cccc|c}
        & \multicolumn{4}{c|}{\textbf{Age}} & \\
        Gender & $\le 18$ & $]18,25]$ & $]25, 40]$ & $> 40$ & Total\\
        \hline
        Woman & 50 & 150 & 100 & 25 & 325 \\
        Man   & 40 & 120 & 80 & 16 & 256\\
        \hline
        Total & 90 & 270 & 180 & 41 & 581
    \end{tabular}
    \caption{Sample stratified according to gender and age group.}
    \label{tab:frequenties-steekproef}
\end{table}

The advantage of a stratified sample is that it is easier to check whether the sample is representative for the whole population. 
Because in this example the researchers made a selection based on specific characteristics, the sample, however, cannot be regarded as random.

After the sample is stratified, one must determine how the required number of objects or respondents must be selected within each stratum. If possible, this should be done in a random manner.

%\subsubsection{Andere steekproefmethoden}

% TODO: Eventueel andere steekproefmethoden opsommen, bv. ahv
% https://www.studiemeesters.nl/scriptie/steekproefmethode-steekproef-nemen-doe-je-zo/

\subsection{Sampling Errors}

No matter what you do to take the sample as good as possible, your measurements will almost always differ from what you would get if you examined the entire population. There will most likely be \emph{errors} in the results. These errors can be subdivided into the following categories:


\paragraph{Accidental Sampling Errors}

Even when using a a random sample, it is possible that measurements within the sample deviate from what would have been measured in the full population. These errors are called \emph{accidental sampling errors}, and this type of error is inevitable.

%Ook als je een aselecte steekproef neemt, is het mogelijk dat bij de geselecteerde elementen de metingen afwijken van wat in de populatie als geheel zou gemeten worden. In dit geval spreken we van een \emph{toevallige steekproeffout}. Dit soort fouten is onvermijdelijk.

\paragraph{Systematic Sampling Errors}

A procedure in the sample that results in an error that has a systematic cause and is therefore not due to accidental effects.
For example, by systematically questioning a privileged part of the population.
If you conduct an online survey, you exclude anyone who does not have an internet connection.
People will also be more likely to participate in a survey if they have a certain affinity with the subject.

\begin{example}
    Scientific literature, including top journals, regularly makes general statements about human psychology and behavior based on samples taken entirely from Western, highly educated, industrialized, rich and democratic societies. These properties (Western, Educated, Industrialized, Rich, and Democratic) are often abbreviated as \emph{WEIRD}.
    
    The scientists assume that these groups are representative for the global population as a whole, but that is a dangerous assumption.
    A study by~\textcite{HenrichEtAl2010} concludes that there are important indications that the members of \emph{WEIRD} societies exhibit unusual qualities in relation to the world population as a whole, and can sometimes even be regarded as outliers!
    
    As a result, conclusions drawn from this type of research will be put at risk!
\end{example}

\paragraph{Accidental Non-Sampling Errors}

A non-sampling error is not related to the way the sample is set up, but rather to the measurement or the collection of information within the sample.

Examples of \emph{accidental non-sampling errors} include incorrectly ticked answers, either by accident or because the respondent interpreted the question differently than the researcher intended.

These types of errors can be avoided, for example by formulating the questions clearly and not misinterpretably, by designing and applying a strict measurement procedure, etc.

\subsubsection{Systematic Non-Sampling Errors}

For example, if respondents with a strong link to the survey are more likely to complete a questionnaire, you will get more positive answers - while they are not representative for the entire population.

\section{Exercises}
\label{sec:onderzoeksproces-oefeningen}

\subsection{Research Process}

In the Bachelor thesis by \textcite{Akin2016}, a comparative study between several persistence methods in Android is conducted. The abstract reads:

\begin{displayquote}
  Nowadays, there are a lot of applications, but which ones remain functional without an internet connection? At present, supporting offline operation of an application is no longer a luxury, but a must-have. In order to provide offline support in an application, a database is used. Because of this, databases are important within the IT industry.
  
  There are different types of databases, but which one to use? Which type is most suitable for a specific type of application? The choice of database can have a major impact on several properties: performance, startup speed, CPU usage, \ldots If the database negatively affects these metrics, this may have as consequence that the number of users of the application will drop.
  
  To find an answer to the problem statement, the following partial questions with regards to the application were formulated:
  
  \begin{itemize}
    \item What is the effect of the chosen database on startup speed? Does the database slow down the startup or does it not have any effect compared to other databases?
    \item What is the effect of the chosen database on CPU usage? A higher CPU usage will lead to higher battery drain. Will the CPU usage be higher or lower compared to other databases?
    \item What is the average speed of the chosen database when adding records to the database?
  \end{itemize}
  
  The research was conducted using three different application profiles: a small amount of data (profile 1), a medium amount of data (profile 2), and a large amount of data (profile 3).
  
  The expectations were that Realm would always have been the better choice, except for application profile 1. In that case, SharedPreferences should be the better choice, since it was specifically designed for small amounts of simple data.
  
  However, experiments yielded the following results:
  
  \begin{enumerate}
    \item Small amount of data: Realm
    \item Medium amount of data: Realm
    \item Large amount of data: SQLite
  \end{enumerate}

\end{displayquote}

\begin{exercise}
  After reading the abstract, try to answer to the following questions as good as possible:
  
  \begin{enumerate}
    \item What is the objective of this research?
    \item Who is the intended target audience?
    \item Are conclusions stated explicitly? 
    \item How is the text structured? Does this match the prescribed structure discussed in class?
  \end{enumerate}
\end{exercise}

\begin{exercise}
  Find the following components of an abstract in the text:
  
  \begin{itemize}
    \item Context
    \item Need
    \item Task
    \item Objective
    \item Result
    \item Conclusion
    \item Perspective
  \end{itemize}

  If you were unable to answer some of the questions above, try to formulate an answer as if you were to conduct this research yourself.
\end{exercise}

\subsection{Basic Concepts in Research}


\begin{exercise}[Retrieval Practice: Measurement Levels]
    \label{ex:retrieval-practice-measurement-levels}
    Measurement levels are an important concept in descriptive statistics because most visualization and analysis techniques depend on this. It is therefore important to know and understand the different measurement levels.
    
    \emph{Retrieval practice} is a study technique that has been scientifically proven to be effective and which leads to improved learning outcomes~\parencite{RoedigerKarpicke2006}.
    
    \begin{enumerate}[label=\alph*.]
        \item Take a blank sheet of paper and try to reproduce an overview of all measurement levels without consulting the syllabus or other sources. Describe the specific properties for each measurement level and give some examples.
        
        \textbf{Take enough time for this} (e.g. at least 5 to 10 minutes). Do not immediately look into the syllabus, but try to remember as much as possible.
        
        When you are done with this, indicate everything that you have noted so far with a (e.g. green) marker.
        
        \item If possible, consult with a fellow student and try to complete the overview that you have each made. Mark all additions with a marker in a different color (e.g. yellow).
        
        \item Finally verify your notes using the syllabus and correct/complete incorrect or missing information if necessary. Indicate this in a third color (e.g. orange or red).
    \end{enumerate}
    
    Thanks to the highlighted colors in your notes, you now have an overview of what you already know and what you need to study. 
    Repeat this exercise a number of times during the course of the semester. 
    Never first look at the result of a previous attempt, but immediately try to retrieve as much information as possible from your memory on a blank sheet. Compare the results afterwards. 
    You should notice that over time, you will mark more and more in green and you will use less red or even none at all.
\end{exercise}

A data frame or tibble in R can represent the result of a \textit{sample}. 
In this data frame, each row represents an \textit{observation}\index{observation} and each column a \textit{variable}\index{variable}. 
You can select a variable using the notation \texttt{dataFrame\$columnName}. 
Observations can be selected using the row number (e.g. \texttt{dataset[1,]}) or using a property (e.g. \verb|dataset[dataset$a == "yes",]|).

\begin{exercise}
    Import the dataset containing the results of~\textcite{Akin2016} in R. 
    You can find this file in the GitHub repository of this course, in the folder
    \texttt{exercises/datasets/}, 
    file \texttt{android\_persistence\_cpu.csv}.
    
    \begin{enumerate}
        \item Get an overview of the different variables used in this dataset.
        \item Use the functions \texttt{glimpse()} and \texttt{View()} to view the contents of the dataset.
        \item Use the function \texttt{unique()} to get a list of distinct values for each variable in the sample.
        \item Define the measurement level for each variable. The results from the previous steps can help!
      \end{enumerate}
\end{exercise}

\begin{exercise}
  Import the dataset \texttt{earthquakes.csv} in R.
  What is the measurement level of the following variables?
  
  \begin{enumerate}
    \item Latitude and Longitude
    \item Type
    \item Time
    \item Depth
  \end{enumerate}

  Note that in this dataset, the column \texttt{ID} is not really a variable. 
  This column was only added to get a unique key for each obvervation in the sample, but does not represent a property.
\end{exercise}

\subsection{Sampling Methods and Sampling Errors}

\begin{exercise}
    A researcher wants to investigate as accurately as possible the consumption habits of residents ages 18 years and older in a certain municipality with 3 residential areas. 
    He distinguishes 4 age groups so he eventually has a total of 12 subgroups. 
    From the municipality, he requests the percentual composition of the population in the city and based on the numbers he calculated the number of respondents required for each age group.
    We call this type of sample a \emph{quota sample}.
    
    Questions:
    \begin{enumerate}[label=\alph*.]
      \item Is this a random sample? Why (not)?
      \item Is this sample representative for the population?
      \item Which types or sampling errors can occur?
      \item What are the advantages and disadvantages?
      \item What other parameters could be used to divide the population in subgroups?
    \end{enumerate}
  \end{exercise}
  
  \begin{exercise}
    A research agency wants to investigate the purchasing behavior of washing products.
    They interrogate a number of women between 25 and 55, because they assume that these are representative for most customers.
    
    \begin{enumerate}[label=\alph*.]
      \item Is this a random sample? Why (not)?
      \item Which sampling error was made?
      \item What is the impact of this error?
    \end{enumerate}
  \end{exercise}
  
  \begin{exercise}

    

    The trade union wants to investigate the working conditions of the employees of an IT company.
    This company has a total of 3200 employees spread over 12 branch offices.
    Because the total number of employees is quite large, 40 employees are selected at random from each branch office.
    The size of the sample is therefore $ n = 480 $.
    
    \begin{enumerate}[label=\alph*.]
      \item Is this a random sample? Why (not)?
      \item Which sampling error was made?
      \item What is the impact of this error?
      \item In what scenario could this sample be representative for the whole population?
    \end{enumerate}
  \end{exercise}
  
  \begin{exercise}
    We want to conduct a study about students at HOGENT, faculty of Business and Organization.
    For this study, the students present during a certain course unit are questioned.
    
    \begin{enumerate}[label=\alph*.]
      \item Is this a random sample? Why (not)?
      \item Which sampling error was made?
      \item Suppose the teacher is very strict in the students' perception and walks around during the survey. What objection can be made here? More specifically, what error can be introduced this way?
      \item Suppose the survey is not taken during a lesson, but after an exam. What error could be made?
    \end{enumerate}
  \end{exercise}
  
  \begin{exercise}[Retrieval practice: sampling errors]
    Use the procedure for retrieval practice from Exercise~\ref{ex:retrieval-practice-measurement-levels} to study the different types of \emph{sampling errors}.
    
    Give an overview of the different types of sampling errors, describe them and give an example.
  \end{exercise}

