---
title: "Exercise 3.8. Formula variance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Why does the formula for the variance use the square of the differences? 
Why not the actual values themselves (i.e. $(\overline{x} - x_i)$) or the absolute value of the difference (i.e. $\left|\overline{x} - x_i\right|$)?

```{r}
variance1 <- function(x) {
  m <- mean(x)
  
  sum(m - x) / (length(x) - 1)
}

variance2 <- function(x) {
  m <- mean(x)
  
  sum(abs(m - x)) / (length(x) - 1)
}

variance3 <- function(x) {
  m <- mean(x)
  
  sum((m-x)^2) / (length(x) - 1)
}
```

Try out each variant of the equation for both datasets $X = \left\{4,4,-4,-4\right\}$ and $Y = \left\{7,1,-6,-2\right\}$.

```{r}
x <- c(4, 4, -4, -4)
y <- c(7, 1, -6, -2)
```

## Results formula 1:

```{r}
variance1(x)
variance1(y)
```

The data is both datasets is obviously spread, but the formula returns the result 0 because the data points are both on the left and right side of the mean and cancel out each other.

## Results formula 2

```{r}
variance2(x)
variance2(y)
```

The dispersion of both datasets is clearly different, min and max are more spread in `y`.


```{r}
range(x)
range(y)
```

The formula however returns 16/3 for both datasets. Therefore, this is not a good measure for the dispersion.

## Results formula 3

```{r}
variance3(x)
variance3(y)
```

For dataset `y`, in which the points are more spread, the result is effectively larger than for dataset `x`.
By taking the squared difference, points that are further away from the mean will increase the result.

### Results using var
We could also use the built-in function `var` to calculate the variance. 

```{r}
var(x)
var(y)
```
