---
title: "Exercise 3.7. The ais dataset"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Remark:** Before running this code in RStudio, first do (from the menu): *Session > Set Working Directory > To Source File Location*. This allows loading the .csv file from the current directory.

## Question

Consider the following subsets of the data frame `ais`:

1. the rowers;
2. the rowers, the netballers and the tennis players combined;
3. the female basketball players and rowers.

## Loading the data

```{r}
# Load data from csv
ais <- read.csv("../../../onderzoekstechnieken-cursus/oefeningen/datasets/ais.csv", sep = ",")
# Allows the columns to be accessed directly instead of with table$column notation
attach(ais)
```

## Rowers

The data set consists of 13 variables, 11 quantitative and 2 qualitative variables.
As an example, we will look at 2 quantitative variables and 1 qualitative variable.

1. Select all rowers from the dataset
    
    ```{r}
    roeiers <- ais[ais$sport == 'Row',]
    ```

2. Calculate the appropriate measures for central tendency and dispersion for the variable `ht` and visualize the results.

    ```{r}
    summary(roeiers$ht)  # min, max, quartiles, mean
    mean(roeiers$ht)     # mean
    sd(roeiers$ht)       # standard deviation of a sample
    min(roeiers$ht)      # smallest value
    median(roeiers$ht)   # median
    IQR(roeiers$ht)      # interquartile range
    max(roeiers$ht)      # largest value
    quantile(roeiers$ht, c(0,0.25,0.5,0.75,1)) # Quantiles
    
    boxplot(roeiers$ht, horizontal = T)        # Boxplot of the dispersion
    ```

3. Calculate the appropriate measures for central tendency and dispersion for the variable `hc` and visualize the results.

    ```{r}
    summary(roeiers$hc)  # min, max, quartiles, mean
    mean(roeiers$hc)     # mean
    sd(roeiers$hc)       # standard deviation of a sample
    min(roeiers$hc)      # smallest value
    median(roeiers$hc)   # median
    IQR(roeiers$hc)      # interquartile range
    max(roeiers$hc)      # largest value
    quantile(roeiers$hc, c(0,0.25,0.5,0.75,1)) # Quantiles
    
    boxplot(roeiers$hc, horizontal = T)        # Boxplot of the dispersion
    ```

4. Calculate the appropriate measures for central tendency and dispersion for the variable `sex` and visualize the results.

    ```{r}
    roeiers_sex_frq <- table(roeiers$sex)       # frequency table
    roeiers_sex_frq[which.max(roeiers_sex_frq)] # mode
    
    barplot(roeiers_sex_frq)    # Barplot of the frequency table
    mosaicplot(roeiers_sex_frq) # Mosaic Plot
    ```

## Rowers, the netballers and the tennis players combined

In this example we will split the values over the three groups.

1. Select all data for the three groups from the dataset:
    
    ```{r}
    sporters <- ais[ais$sport == 'Row' |
                    ais$sport == 'Netball' |
                    ais$sport == 'Tennis',]
    # The following function is used to omit sports that are not
    # included in the subset in e.g. plots.
    sporters$sport <- factor(sporters$sport)
    View(sporters)
    ```

2. Calculate the appropriate measures for central tendency and dispersion for the variable `ht` and visualize the results

    ```{r}
    # min, max, quartiles, mean
    aggregate(ht ~ sport, data = sporters, FUN = summary)
    # mean
    aggregate(ht ~ sport, data = sporters, FUN = mean)
    # standard deviation of a sample
    aggregate(ht ~ sport, data = sporters, FUN = sd)
    # median
    aggregate(ht ~ sport, data = sporters, FUN = median)
    # interquartile range
    aggregate(ht ~ sport, data = sporters, FUN = IQR)
    # quantiles
    aggregate(ht ~ sport, data = sporters, FUN = quantile)

    # Boxplot of the dispersion
    boxplot(sporters$ht ~ sporters$sport, horizontal = T)       
    ```

3. Analysis of the variable `sex`

    ```{r}
    # Frequency table for distribution of m/f for each sport
    ft <- table(sporters$sex, 
                sporters$sport)
    ft
    # Mosaic Plot
    mosaicplot(sporters$sport ~ sporters$sex,
               main = "Distribution of m/f for each sport",
               xlab = "sport", ylab = "sex")
    # Bar Plot
    barplot(ft, legend = TRUE,
            main = "Distribution of m/f for each sport",
            xlab = "sport", ylab = "sex")
    ```

## The female basketball players and rowers.

1. Select all data from the dataset
   
    ```{r}
    vrouwen <- ais[ais$sex == 'f' & 
                     (ais$sport == 'B_Ball' | 
                      ais$sport == 'Row'),]
    vrouwen$sex <- factor(vrouwen$sex)
    vrouwen$sport <- factor(vrouwen$sport)
    View(vrouwen)
    ```
    
2. Calculate the appropriate measures for central tendency and dispersion for the variable `ht` and visualize the results

    ```{r}
    # min, max, quartiles, mean
    aggregate(ht ~ sport, data = vrouwen, FUN = summary)
    # mean
    aggregate(ht ~ sport, data = vrouwen, FUN = mean)
    # standard deviation of a sample
    aggregate(ht ~ sport, data = vrouwen, FUN = sd)
    # quantiles
    aggregate(ht ~ sport, data = vrouwen, FUN = quantile)

    # Boxplot of the dispersion
    boxplot(vrouwen$ht ~ vrouwen$sport, horizontal = T)       
    ```

3. Analysis of the variable `sport`

    ```{r}
    # Frequency table
    ft_2 <- table(vrouwen$sport)
    ft_2
    # Bar Plot
    barplot(ft_2, legend = TRUE)
    ```

