---
title: "Exercise 3.11. Mean and Range"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question

Use the functions `mean` and `range` in R to determine the mean and range of:

1. The numbers 1,2, ..., 21
2. 50 random values, generated using a normal distribution with mean 0 and variance 1 (function `rnorm`)
3. Columns `height` and `weight` in the data frame `women` (default in R)


## Oplossing

1. The numbers 1,2, ..., 21

    ```{r}
    # Create a list containing these numbers and calculate mean and range
    numbers <- 1:21
    mean(numbers)
    range(numbers)
    ```

2. 50 random values, generated using a normal distribution with mean 0 and variance 1

    ```{r}
    # Generate random values, parameters can be omitted (default values)
    numbers2 <- rnorm(50, mean = 0, sd = 1)
    mean(numbers2)
    range(numbers2)
    ```

3. Columns `height` and `weight` in the data frame `women`

    ```{r}
    mean(women$height)
    range(women$height)
    mean(women$weight)
    range(women$weight)
    ```